---
title: "Darwin Core mapping"
subtitle: "For: my_dataset_title"
author:
- Lissa Breugelmans
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c("tidyverse", "tidylog", "magrittr", "here", "janitor", "readxl", "digest", "rgbif")
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(readxl)         # To read Excel files
library(digest)         # To generate hashes
library(rgbif)          # To use GBIF services
```

# Read source data

Create a data frame `raw`_data` from the source data:

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/e/2PACX-1vSQE6xFj-Lbaw-J5fwNNFoKWbg2jHiKMx9Fzbofc4RihnBCJE4oL-wAnkTDHorS2kYC4dpMEgaRKWAS/pub?gid=1172557183&single=true&output=csv).

Read the relevant worksheet (published as csv):

```{r read_source_data}
raw_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSQE6xFj-Lbaw-J5fwNNFoKWbg2jHiKMx9Fzbofc4RihnBCJE4oL-wAnkTDHorS2kYC4dpMEgaRKWAS/pub?gid=1172557183&single=true&output=csv", col_types = cols(.default = "c"))
```

# Process source data

## Tidy data

Clean data somewhat:

```{r}
input_data <-
  raw_data |>
  remove_empty("rows") |>
  clean_names()
```

Save the source data to the repository to keep track of changes:

```{r}
write_csv(raw_data, here("data", "raw", "raw_data.csv"), na = "")
```

# Taxon core

## Term mapping

### Record-level terms

Create a dataframe with unique taxa only (ignoring multiple distribution rows). Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records).

```{r}
taxon <-
  input_data |>
  distinct(taxon_key, .keep_all = TRUE) |>
  mutate(
    language = "en", # e.g. "en"
    license = "https://creativecommons.org/licenses/by/4.0/",
    rightsHolder = "APM", 
    datasetID = "my_dataset_doi",
    institutionCode = "APM",
    datasetName = "Checklist of Bryophytes in Belgium"
  )
```
### Taxon-level terms

The following terms contain information about the taxon:

```{r}
taxon <-
  taxon |>
  dplyr::mutate(
    taxonID = taxon_key,
    scientificName = scientific_name,
    kingdom = kingdom,
    phylum = phylum,
    class = class,
    order = order,
    family = family,
    genus = genus,
    taxonRank = taxon_rank,
    nomenclaturalCode = nomenclatural_code
  )
```

## Post-processing

Arrange by taxonID and only keep the Darwin Core columns:

```{r}
taxon <-
  taxon |> 
  arrange(taxonID) |> 
  select(
    "language", "license", "rightsHolder", "datasetID", "institutionCode",
    "datasetName", "taxonID", "scientificName", "kingdom", "phylum", "class",
    "order", "family", "genus",
    "taxonRank", "nomenclaturalCode"
  )
```



# Distribution extension

## Pre-processing

Create a dataframe with all data:

```{r}
distribution <- input_data
```

## Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id)
```

### locality

Inspect values:

```{r}
distribution %>%
  group_by(country_code, locality) %>%
  count()
```

Map values to `locality` if provided, otherwise use the country name:

```{r}
distribution %<>% mutate(dwc_locality = case_when(
  !is.na(locality) ~ locality,
  country_code == "BE" ~ "Belgium",
  country_code == "GB" ~ "United Kingdom",
  country_code == "MK" ~ "Macedonia",
  country_code == "NL" ~ "The Netherlands",
  TRUE ~ "" # In other cases leave empty
))
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(country_code, locality, dwc_locality) %>%
  count()
```

### countryCode

Inspect values:

```{r}
distribution %>%
  group_by(country_code) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_countryCode = country_code) 
```

### occurrenceStatus

Inspect values:

```{r}
distribution %>%
  group_by(occurrence_status) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceStatus = occurrence_status) 
```

### threatStatus

Inspect values:

```{r}
distribution %>%
  group_by(threat_status) %>%
  count()
```

Map values by recoding to the [IUCN threat status vocabulary](http://rs.gbif.org/vocabulary/iucn/threat_status.xml):

```{r}
distribution %<>% mutate(dwc_threatStatus = recode(threat_status,
  "endangered" = "EN",
  "vulnerable" = "VU"
))
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(threat_status, dwc_threatStatus) %>%
  count()
```

### source

Inspect values:

```{r}
distribution %>%
  group_by(source) %>%
  count() %>%
  head() # Remove to show all values
```

Map values:

```{r}
distribution %<>% mutate(dwc_source = source) 
```

### occurrenceRemarks

Inspect values:

```{r}
distribution %>%
  group_by(remarks) %>%
  count() %>%
  head() # Remove to show all values
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceRemarks = remarks) 
```

## Post-processing

Only keep the Darwin Core columns:

```{r}
distribution %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
```

Preview data:

```{r}
distribution %>% head()
```

Save to CSV:

```{r}
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
```
